---
title: "Practice session for Econometrics for Public Policy"
date: "April 9, 2025"
author: "Donnie Osborne"
fontsize: 10pt
output:
  beamer_presentation:
    theme: Singapore
  slidy_presentation: default
  ioslides_presentation: default
header-includes:
- \usepackage{setspace}
- \setstretch{1.25}
- \setbeamertemplate{navigation symbols}{}
- \setbeamertemplate{footline}[page number]
---

# Introduction

## Introduction

-   Donnie Osborne
-   Affiliation: Graduate School of Economics, PhD
-   Research area: Development Economics
-   Office hours: By appointment
    -   Contact by email: [\textcolor{blue}{osborne-donnie@g.ecc.u-tokyo.ac.jp}](mailto:osborne-donnie@g.ecc.u-tokyo.ac.jp){.email}

## Course Information

-   This course is a companion course of Econometrics for Public Policy by Professor Chen
    -   Delivery type: Hybrid
    -   Course time: Wed, period 3 (13:00 - 14:45)
-   [\textcolor{blue}{Zoom meeting}](https://u-tokyo-ac-jp.zoom.us/j/3536507388?pwd=bDlXandsWVUyWEdjNXUyS0hZeE5BZz09) information:
    -   Meeting ID: 353 650 7388
-   GraSPP students who register for Econometrics for Public Policy **are required** to take this course

## Course Information

-   The goal of this course is to teach you how to use R to conduct empirical research
-   Basically, we need **empirical toolkits**, **data**, and **econometrics software** to conduct empirical research
-   In this course, you will learn how to do data cleaning work and conduct applied economic research using R
    -   Basic commands
      -   dplyr, tidyr, ggplot2
    -   Advanced packages for causal analysis
      -   estimatr, rdd, fixest, didimputation, did
-   You can find the course materials in this [\textcolor{blue}{google drive}](https://drive.google.com/drive/folders/1QDYXm2Rw3eVAUK2MpiY9dOjMr558L5MR?usp=drive_link)
-   Please use your ECCS Cloud Mail to get access

## Course Logistics

-   Professor Chen will teach the theoretical part of this course, so I will focus on the **R implementation**

-   Basically, this course is divided into two parts:

    -   Basic techniques for cleaning, analyzing, and plotting data
    -   Empirical commands for causal inferences

## Homework and Grading

-   There are six homeworks this semester

    1.  HW1 (4/16 - 5/7)
    2.  HW2 (5/7 - 5/21)
    3.  HW3 (5/21 - 6/4)
    4.  HW4 (6/4 - 6/18)
    5.  HW5 (6/18 - 7/2)
    6.  HW6 (7/2 - 7/15)

-   Please submit your homework to the grader by email

    -   Grader: Jovey Mae DOMINGO
    -   Email: [\textcolor{blue}{grasppmetrics2025@gmail.com}](mailto:grasppmetrics2025@gmail.com){.email}
    -   If you have any problems with the homework, please contact the grader using the email above

## Homework and Grading

-   Please submit both the R script or Markdown file and compiled report or knit file
-   Late submissions will be discounted by 50% per day 
-   You can discuss homework with your classmates, but each individual must submit
-   chatGPT: use it as you see fit; however, the code submitted must reflect the lectures
-   **You won't get credit just because your code runs the correct output!**
-   Course Evaluation:
    -   HW1 through HW5 = 15% each
    -   HW6 = 25%

## Outline for Today

-   Introduction to R
-   Access to R
-   Basics of R
    -   Common functions
    -   Interface
    -   Type commands
    -   Typical format
    -   Some basic commands
-   Homework 0

# Introduction to R

## Installation and setup of R

1.  Install Rtools for Windows machines from [\textcolor{blue}{CRAN Rtools}](https://cran.r-project.org/bin/windows/Rtools/).

2.  Install the latest version of R from [\textcolor{blue}{CRAN R}](https://cran.r-project.org).

3.  Install the latest version of [\textcolor{blue}{RStudio}](https://www.rstudio.com/products/RStudio/). RStudio is the editor we’ll rely on, i.e. we’ll write code in RStudio which is subsequently sent to and run within R.

## Introduction to R

-   What is R?
    -   R is a powerful programming language and software environment designed for *data analysis, statistics, and visualization*. It is widely used by *scientists, researchers, and data analysts* to make sense of large amounts of data.

-   Why Learn R?
    -   Free and Open Source
    -   Great for Data Analysis
    -   Strong Visualization Tools
    -   Widely Used in Industry

## What can you do with R?

-   Data management
    -   Importing and exporting data in various formats
    -   Cleaning and transforming data
    -   Merging and appending data sets
-   Data visualization
    -   Scatter plot, histograms, PDFs and CDFs
    -   Non-parametric plots
    -   Regression discontinuity and dynamic DiD
-   Data analysis
    -   Linear regression
    -   Panel data analysis
    -   Time series analysis
    -   Causal inference and structural analysis

## Example: Working hours per worker

-   The data set provides estimates of average annual working hours per worker across countries between 1950 and 2017 ([\textcolor{blue}{Giattino, Ortiz-Ospina, and Roser, 2020}](https://ourworldindata.org/working-hours))

-   We will use this example to illustrate some R functions

    -   Data management
        -   Import an Excel file into R
        -   Create a new variable for weekly working hours per worker
    -   Data visualization
        -   Plot a time-series graph
    -   Data analysis
        -   Use linear regression to examine whether workers in East Asia work longer

# RStudio

## R's Interface

1.  *Source Pane* is a text editor that will be referred to as R Script
    -   You can write commands and save them, which is the main point of reproducibility

2.  *Console Pane* is the place you write your line-by-line command 
    -   It means you can only write a single command or a long paragraph of commands 
    -   After you close the RStudio, the commands will not be saved unless you specified to do so 
    -   Example: Symbol \> called prompt
        -   Type 3 + 4, and press Enter

3.  *Environment Pane* has several sub-windows 

    -   Global Environment is the place where your data will be after importing data
    -   History saves the commands you run in R console
    -   Connections is where you connect to external databases
  
## R's Interface

4.  *Files Pane* also has several sub-windows

    -   You can manage files and folders as well as set the working directory
    -   Plots is where your plots will appear
    -   Packages is where you manage your R packages - you can install it from CRAN or other repositories
    -   Help is where R stores documentations - you can open the help or introduction page of the respective packages as well as individual functions
      -   Example: ?lm() (in stock package 'stats')

## R Scripts and R Markdown

1.  R Scripts

    -   Rather than typing everything into the console, this allows you to type and save your commands.
    -   This can be manually edited with comments.

2.  R Markdown

    -   A versatile format for analysis and tracking progression.
    -   Helpful for reproducibility.
    -   Can be used to create various dynamic output files.
    
# R Basics    
    
## Basics: packages and libraries

\footnotesize

**Packages** only have to be loaded once

```{r, size = "tiny", echo = TRUE, eval = FALSE, results = 'hide'}
install.packages("tidyverse") # data transformation functions
```

**Libraries** have to be loaded each session

```{r, size = "tiny", echo = TRUE, eval = FALSE}
library(tidyr)
library(dplyr)
```

## Basics: operators

\footnotesize

**Assignment**

```{r, size = "tiny", echo = TRUE, eval = FALSE}
x <- 1
y <- 2
```

**Arithmetic**

```{r, size = "tiny", echo = TRUE, eval = FALSE}
x+y
x*y
```

## Basics: operators

\footnotesize

**Comparison**

```{r, size = "tiny", echo = TRUE, eval = FALSE}
x==y # equal
x!=y # not equal
x>y # greater than
x<=y # less than or equal to
```

**Logical**

```{r, size = "tiny", echo = TRUE, eval = FALSE}
y>x && y==2*x # AND: both conditions must be true
x > y || x < y # OR: either condition must be true
```

## Basics: data types

\footnotesize

**Numeric**

*Double*: decimal numbers

```{r, size = "tiny", echo = TRUE, eval = FALSE}
x <- 1.5
typeof(x) # returns object type
is.numeric(x) # returns if object is numeric or not
```
*Integer*: whole numbers

```{r, size = "tiny", echo = TRUE, eval = FALSE}
x <- 1L # L is necessary to make it an integer
```

**Character**

```{r, size = "tiny", echo = TRUE, eval = FALSE}
x <- "Hello"
```

```{r, size = "tiny", echo = TRUE, eval = FALSE}
x <- "5"
```

*Perform an arithmetic operation on x now*

## Basics: data structure

\footnotesize

**Vectors**: all elements must be of the same type

```{r, size = "tiny", echo = TRUE, eval = FALSE}
x <- sample(1:10, 5, replace = TRUE)
y <- sample(1:10, 5, replace = TRUE)
year <- 1990:1994
names <- c("payal", "amazon", "google", "microsoft", "alibaba")
```

**Lists**: can contain elements of different types and lengths

```{r, size = "tiny", echo = TRUE, eval = FALSE}
people <- list(name = c("bob", "jan", "mitsuha"), 
               age = c(24, 36, 21), 
               grade = c("1st", "2nd", "4th"))
```

**Accessing** elements in the list

```{r, size = "tiny", echo = TRUE, eval = FALSE}
people[1]
people[[1]]
people[[1]][1]
```

## Basics: data structure

**Exercise**: 

1. Create the following vectors
    -   classes = classes you are taking
    -   period = period of the corresponding classes
    -   program = program you are in
    -   year = year you are in
    
2. Store these vectors in a list

3. Pull from the list the period of this class (output should = 3)


## Basics: data structure

\footnotesize

**Data Frames**: two-dimensional of equal length (list of vectors)

```{r, size = "tiny", echo = TRUE, eval = FALSE}
df <- data.frame(names, year, x, y) # combines vectors into data frame
tibble(df) # see the data frame
```

**Accessing** the vectors (*variables*) in the data frame

```{r, size = "tiny", echo = TRUE, eval = FALSE}
df$year
```

```{r, size = "tiny", echo = TRUE, eval = FALSE}
df %>% pull(year) # piping: lecture 2
```

# Homework 0

## Homework 0: Set up your own Script/RMarkdown

\footnotesize

1.  Load the library readxl

```{r, size = "footnotesize", echo = TRUE, eval = FALSE}
install.packages("readxl") 
library(readxl)
```

2.  Import the data from the excel file and store it in the global environment

```{r echo = TRUE, eval = FALSE, size = "small"}
setwd("C://Users//...") # two forward slashes instead of back slash
data <- read_excel(file.path("annual-working-hours-per-worker.xlsx"), 
                   sheet = "annual-working-hours-per-worker")
```

3.  Skim the data with tibble: what are the data types of the variables?

```{r echo = TRUE, eval = FALSE, size = "small"}
tibble(data)
```

4.  Render the file

```{r echo = TRUE, eval = FALSE, size = "small"}
rmarkdown::render("file_name.R", "pdf_document")
```
  